Variational Monte Carlo on bosonic systems

### 2018-02-06 project_1/article/article.tex
Update derivation of the full problem

\section{Theory}
    To model the trapped bosonic gas particles we use the potential
    \begin{align}
        V_{\text{ext}}(\vf{r})
        &=
        \begin{cases}
            \half m\omega^2r^2 & (\text{S}), \\
            \half m \bigl[
                \omega^2(x^2 + y^2) + \omega_z^2z^2
            \bigr] & (\text{E}),
        \end{cases}
    \end{align}
    where we can choose between a spherical (S) or an elliptical (E) harmonic
    trap. The two-body Hamiltonian of the system is given by
    \begin{align}
        H = \sum_{i}^{N}h(\vf{r}_i) + \sum_{i < j}^{N}w(\vf{r}_i, \vf{r}_j),
    \end{align}
    where the single particle one body operator, $h$, is given by
    \begin{align}
        h(\vf{r}_i) = -\frac{\hbar^2}{2m}\nabla_i^2
        + V_{\text{ext}}(\vf{r}_i),
    \end{align}
    (we assume equal mass)
    and the two-body interaction operator, $w$, is
    \begin{align}
        w(\vf{r}_i, \vf{r}_j)
        = \begin{cases}
            \infty & |\vf{r}_i - \vf{r}_j| \leq a, \\
            0 & |\vf{r}_i - \vf{r}_j| > a,
        \end{cases}
    \end{align}
    where $a$ is the hard sphere of the particle. The trial wavefunction,
    $\ket{\Psi_T}$, we will
    be looking at is given by
    \begin{align}
        \Psi_T(\vf{r})
        &= \Phi_T(\vf{r})
        \prod_{j < k}^N f(a, \vf{r}_j, \vf{r}_k) \\
        &= \Biggl(
            \prod_i^N g(\alpha, \beta, \vf{r}_i)
        \Biggr)
        \prod_{j < k}^N f(a, \vf{r}_j, \vf{r}_k),
        \label{eq:initial_trial_wavefunction}
    \end{align}
    where $\alpha$ and $\beta$ are variational parameters and
    \begin{align}
        \vf{r} = (\vf{r}_1, \vf{r}_2, \dots, \vf{r}_N, \alpha, \beta).
    \end{align}
    Here $g$ are the single particle wavefunctions given by
    \begin{align}
        g(\alpha, \beta, \vf{r}_i)
        = \exp\bigl[
            -\alpha(x_i^2 + y_i^2 + \beta z_i^2)
        \bigr] \equiv \phi(\vf{r}_i),
    \end{align}
    and $\ket{\Phi_T}$ the \textit{Slater permanent} consisting of the $N$ first
    single particle wavefunctions, and the correlation wavefunction, $f$, given
    by
    \begin{align}
        f(a, \vf{r}_j, \vf{r}_k)
        &=
        \begin{cases}
            0 & |\vf{r}_j - \vf{r}_k| \leq a, \\
            \Bigl(
                1 - \frac{a}{|\vf{r}_j - \vf{r}_k|}
            \Bigr) & |\vf{r}_j - \vf{r}_k| > a.
        \end{cases}
    \end{align}

@@

                \bra{\vf{r}}\ket{\Psi_T}

@@

                \bra{\vf{r}}\ket{\Phi_T}

@@

@@

                \nabla_k\bra{\vf{r}}\ket{\Psi_T}
                &= \nabla_k\Bigl[
                    \bra{\vf{r}}\ket{\Phi_T}

@@

                + \bra{\vf{r}}\ket{\Phi_T}

@@

            The gradient of the Slater permanent gives

@@

                \nabla_k\phi(\vf{r}_k)
                \prod_{i \neq k}^N\phi(\vf{r}_i) \\
                &= \frac{\nabla_k\phi(\vf{r}_k)}{\phi(\vf{r}_k)}
                \bra{\vf{r}}\ket{\Phi_T}.

@@

                    \nabla_k\sum_{m < n}^N u(r_{mn}) \\

@@

                        \sum_{m = 1}^{k - 1}
                        \nabla_k u(r_{mk})
                        \sum_{n = k + 1}^N
                        \nabla_k u(r_{kn})
                    \sum_{m \neq k}^N\nabla_k u(r_{km}),

@@

                as $r_{ij} = r_{ji}$.  We are thus left with

@@

                    \nabla_k\bra{\vf{r}}\ket{\Psi_T}
                    \frac{\nabla_k\phi(\vf{r}_k)}{\phi(\vf{r}_k)}
                    \bra{\vf{r}}\ket{\Psi_T}
                    + \bra{\vf{r}}\ket{\Psi_T}
                    \sum_{m \neq k}^N\nabla_k u(r_{km}) \\

@@

                        \frac{\nabla_k\phi(\vf{r}_k)}{\phi(\vf{r}_k)}
                        \sum_{m \neq k}^N\nabla_k u(r_{km})

@@

                    \bra{\vf{r}}\ket{\Psi_T}.

@@

                We can now find the Laplacian of the trial wavefunction. This
                gives

@@

                    \nabla_k^2\bra{\vf{r}}\ket{\Psi_T}
                    &= \nabla_k\nabla_k\bra{\vf{r}}\ket{\Psi_T}
                    = \nabla_k
                        \frac{\nabla_k\phi(\vf{r}_k)}{\phi(\vf{r}_k)}
                        \sum_{m \neq k}^N\nabla_k u(r_{km})
                    \bra{\vf{r}}\ket{\Psi_T} \\

@@

                        \frac{\nabla_k^2\phi(\vf{r}_k)}{\phi(\vf{r}_k)}
                        + \left[
                            \nabla\frac{1}{\phi(\vf{r}_k)}
                        \right]
                        \nabla_k\phi(\vf{r}_k)

@@

                        \sum_{m \neq k}^N\nabla_k^2 u(r_{km})
                    \bra{\vf{r}}\ket{\Psi_T}

@@

                        \frac{\nabla_k\phi(\vf{r}_k)}{\phi(\vf{r}_k)}
                        \sum_{m \neq k}^N\nabla_k u(r_{km})
                    \nabla_k\bra{\vf{r}}\ket{\Psi_T} \\

@@

                        \frac{\nabla_k^2\phi(\vf{r}_k)}{\phi(\vf{r}_k)}
                        - \left[
                            \frac{\nabla_k\phi(\vf{r}_k)}{\phi(\vf{r}_k)}
                        \right]^2

@@

                        \sum_{m \neq k}^N\nabla_k^2 u(r_{km})

@@

                    \bra{\vf{r}}\ket{\Psi_T}

@@

                        \frac{\nabla_k\phi(\vf{r}_k)}{\phi(\vf{r}_k)}
                        \sum_{m \neq k}^N\nabla_k u(r_{km})
                    \bra{\vf{r}}\ket{\Psi_T}.
                We now divide by the trial wavefunction. This simplifies the
                calculations and is more similar to the expression for the local
                energy.
                    \frac{\nabla_k^2\phi(\vf{r}_k)}{\phi(\vf{r}_k)}
                    - \left[
                        \frac{\nabla_k\phi(\vf{r}_k)}{\phi(\vf{r}_k)}
                    \right]^2
                    \sum_{m \neq k}^N\nabla_k^2 u(r_{km})
                        \frac{\nabla_k\phi(\vf{r}_k)}{\phi(\vf{r}_k)}
                        \sum_{m \neq k}^N\nabla_k u(r_{km})

@@

                    \frac{\nabla_k^2\phi(\vf{r}_k)}{\phi(\vf{r}_k)}
                    \sum_{m \neq k}^N\nabla_k^2 u(r_{km})
                    \frac{2\nabla_k\phi(\vf{r}_k)}{\phi(\vf{r}_k)}
                    \sum_{m \neq k}^N\nabla_k u(r_{km})
                        \sum_{m \neq k}^N\nabla_k u(r_{km})

@@

@@

            the single particle functions, $\phi(\vf{r}_k)$, and the interaction
            functions $u(r_{km})$. For the single particle functions we use

@@

                \nabla_k\phi(\vf{r}_k)

@@

                \phi(\vf{r}_k),

@@

            note that the subscripts on the unit vectors $\vf{e}_i$ are

@@

                \nabla_k^2\phi(\vf{r}_k)

@@

                \phi(\vf{r}_k)

@@

                \phi(\vf{r}_k).

>>>>

\section{Theory}
    To model the trapped bosonic gas particles we use the potential
    \begin{align}
        V_{\text{ext}}(\vf{r})
        &=
        \begin{cases}
            \half m\omega^2r^2 & (\text{S}), \\
            \half m \bigl[
                \omega^2(x^2 + y^2) + \omega_z^2z^2
            \bigr] & (\text{E}),
        \end{cases}
    \end{align}
    where we can choose between a spherical (S) or an elliptical (E) harmonic
    trap. The two-body Hamiltonian of the system is given by
    \begin{align}
        H = \sum_{i}^{N}h(\vf{r}_i) + \sum_{i < j}^{N}w(\vf{r}_i, \vf{r}_j),
    \end{align}
    where the single particle one body operator, $h$, is given by
    \begin{align}
        h(\vf{r}_i) = -\frac{\hbar^2}{2m}\nabla_i^2
        + V_{\text{ext}}(\vf{r}_i),
    \end{align}
    (we assume equal mass)
    and the two-body interaction operator, $w$, is
    \begin{align}
        w(\vf{r}_i, \vf{r}_j)
        = \begin{cases}
            \infty & |\vf{r}_i - \vf{r}_j| \leq a, \\
            0 & |\vf{r}_i - \vf{r}_j| > a,
        \end{cases}
    \end{align}
    where $a$ is the hard sphere of the particle. The trial wavefunction,
    $\ket{\Psi_T}$, we will
    be looking at is given by
    \begin{align}
        \Psi_T(\vf{r})
        &= \Phi_T(\vf{r})
        \prod_{j < k}^N f(a, \vf{r}_j, \vf{r}_k) \\
        &= \Biggl(
            \prod_i^N g(\alpha, \beta, \vf{r}_i)
        \Biggr)
        \prod_{j < k}^N f(a, \vf{r}_j, \vf{r}_k),
        \label{eq:initial_trial_wavefunction}
    \end{align}
    where $\alpha$ and $\beta$ are variational parameters and
    \begin{align}
        \vf{r} = (\vf{r}_1, \vf{r}_2, \dots, \vf{r}_N, \alpha, \beta).
    \end{align}
    Here $g$ are the single particle wavefunctions given by
    \begin{align}
        g(\alpha, \beta, \vf{r}_i)
        = \exp\bigl[
            -\alpha(x_i^2 + y_i^2 + \beta z_i^2)
        \bigr] \equiv \phi(\vf{r}_i),
    \end{align}
    and $\ket{\Phi_T}$ the \textit{Slater permanent} consisting of the $N$ first
    single particle wavefunctions, and the correlation wavefunction, $f$, given
    by
    \begin{align}
        f(a, \vf{r}_j, \vf{r}_k)
        &=
        \begin{cases}
            0 & |\vf{r}_j - \vf{r}_k| \leq a, \\
            \Bigl(
                1 - \frac{a}{|\vf{r}_j - \vf{r}_k|}
            \Bigr) & |\vf{r}_j - \vf{r}_k| > a.
        \end{cases}
    \end{align}
    We will for brevity use the notation $\phi(\vf{r}_i) = \phi_i$ and $r_{jk} =
    |\vf{r}_j - \vf{r}_k|$.

@@

                \Psi_T(\vf{r})

@@

                \Phi_T(\vf{r})

@@

            To further shorten the notation we will use $u_{jk} = u(r_{jk})$.

@@

                \nabla_k\Psi_T(\vf{r})

@@

                + \Phi_T(\vf{r})

@@

            The gradient of the Slater permament for particle $k$ is given by

@@

                \nabla_k\phi_k
                \prod_{i \neq k}^N\phi_i
                = \frac{\nabla_k\phi_k}{\phi_k}
                \Phi_T(\vf{r}).

@@

                \nabla_k\sum_{m < n}^N u_{mn} \\
                &= J(\vf{r})
                    \sum_{m = 1}^{k - 1}\nabla_k u_{mk}
                    \sum_{n = k + 1}^N\nabla_k u_{kn}

@@

                \sum_{m \neq k}^N\nabla_k u_{km},

@@

            where the gradient of the interaction term splits the antisymmetric
            sum into two parts. As $r_{ij} = r_{ji}$ we can combine these sums
            into a single sum. This in total yields the gradient

@@

                \nabla_k\Psi_T(\vf{r})

@@

                    \frac{\nabla_k\phi_k}{\phi_k}
                    + \sum_{m \neq k}^N
                    \nabla_k u_{km}

@@

                \Psi_T(\vf{r}).
                \label{eq:gradient_full_wavefunction}

@@

            We now find the Laplacian of the trial wavefunction by taking the
            gradient of \autoref{eq:gradient_full_wavefunction}.

@@

                \nabla_k^2\Psi_T(\vf{r})

@@

                    \nabla_k\Biggl[
                        \frac{\nabla_k\phi_k}{\phi_k}

@@

                    \sum_{m \neq k}^N \nabla_k^2 u_{km}
                \Biggr)\Psi_T(\vf{r})

@@

                    \frac{\nabla_k\phi_k}{\phi_k}
                    + \sum_{m \neq k}^N
                    \nabla_k u_{km}
                \Psi_T(\vf{r}),
            where the squared term came from taking the gradient of the trial
            wavefunction.  To further simplify we divide by the trial
            wavefunction. This yields
                    \frac{\nabla_k^2\Psi_T(\vf{r})}{\Psi_T(\vf{r})}

@@

                        \nabla_k\Biggl[
                            \frac{\nabla_k\phi_k}{\phi_k}

@@

                        \sum_{m \neq k}^N \nabla_k^2 u_{km}

@@

@@

                        \frac{\nabla_k\phi_k}{\phi_k}
                        + \sum_{m \neq k}^N
                        \nabla_k u_{km}

@@

                    \frac{\nabla_k^2\phi_k}{\phi_k}
                    + 2\frac{\nabla_k\phi_k}{\phi_k}
                    \sum_{m \neq k}\nabla_k u_{km}
                    + \sum_{m\neq k}^N\nabla_k^2 u_{km}
                    + \Biggl(
                        \sum_{m \neq k}^N\nabla_k u_{km}
                    \Biggr)^2.

@@

            %\begin{widetext}
            %    We can now find the Laplacian of the trial wavefunction. This
            %    gives
            %    \begin{align}
            %        \nabla_k^2\bra{\vf{r}}\ket{\Psi_T}
            %        &= \nabla_k\nabla_k\bra{\vf{r}}\ket{\Psi_T}
            %        = \nabla_k
            %        \Biggl(
            %            \frac{\nabla_k\phi(\vf{r}_k)}{\phi(\vf{r}_k)}
            %            +
            %            \sum_{m \neq k}^N\nabla_k u(r_{km})
            %        \Biggr)
            %        \bra{\vf{r}}\ket{\Psi_T} \\
            %        &=
            %        \Biggl(
            %            \frac{\nabla_k^2\phi(\vf{r}_k)}{\phi(\vf{r}_k)}
            %            + \left[
            %                \nabla\frac{1}{\phi(\vf{r}_k)}
            %            \right]
            %            \nabla_k\phi(\vf{r}_k)
            %            +
            %            \sum_{m \neq k}^N\nabla_k^2 u(r_{km})
            %        \Biggr)
            %        \bra{\vf{r}}\ket{\Psi_T}
            %        \\
            %        &\qquad
            %        +
            %        \Biggl(
            %            \frac{\nabla_k\phi(\vf{r}_k)}{\phi(\vf{r}_k)}
            %            +
            %            \sum_{m \neq k}^N\nabla_k u(r_{km})
            %        \Biggr)
            %        \nabla_k\bra{\vf{r}}\ket{\Psi_T} \\
            %        &=
            %        \Biggl(
            %            \frac{\nabla_k^2\phi(\vf{r}_k)}{\phi(\vf{r}_k)}
            %            - \left[
            %                \frac{\nabla_k\phi(\vf{r}_k)}{\phi(\vf{r}_k)}
            %            \right]^2
            %            +
            %            \sum_{m \neq k}^N\nabla_k^2 u(r_{km})
            %        \Biggr)
            %        \bra{\vf{r}}\ket{\Psi_T}
            %        \\
            %        &\qquad
            %        +
            %        \Biggl(
            %            \frac{\nabla_k\phi(\vf{r}_k)}{\phi(\vf{r}_k)}
            %            +
            %            \sum_{m \neq k}^N\nabla_k u(r_{km})
            %        \Biggr)^2
            %        \bra{\vf{r}}\ket{\Psi_T}.
            %    \end{align}
            %    We now divide by the trial wavefunction. This simplifies the
            %    calculations and is more similar to the expression for the local
            %    energy.
            %    \begin{align}
            %        \frac{\nabla_k^2\bra{\vf{r}}\ket{\Psi_T}}
            %        {\bra{\vf{r}}\ket{\Psi_T}}
            %        &=
            %        \frac{\nabla_k^2\phi(\vf{r}_k)}{\phi(\vf{r}_k)}
            %        - \left[
            %            \frac{\nabla_k\phi(\vf{r}_k)}{\phi(\vf{r}_k)}
            %        \right]^2
            %        +
            %        \sum_{m \neq k}^N\nabla_k^2 u(r_{km})
            %        +
            %        \Biggl(
            %            \frac{\nabla_k\phi(\vf{r}_k)}{\phi(\vf{r}_k)}
            %            +
            %            \sum_{m \neq k}^N\nabla_k u(r_{km})
            %        \Biggr)^2 \\
            %        &=
            %        \frac{\nabla_k^2\phi(\vf{r}_k)}{\phi(\vf{r}_k)}
            %        +
            %        \sum_{m \neq k}^N\nabla_k^2 u(r_{km})
            %        +
            %        \frac{2\nabla_k\phi(\vf{r}_k)}{\phi(\vf{r}_k)}
            %        \sum_{m \neq k}^N\nabla_k u(r_{km})
            %        +
            %        \Biggl(
            %            \sum_{m \neq k}^N\nabla_k u(r_{km})
            %        \Biggr)^2
            %    \end{align}
            %\end{widetext}

@@

            the single particle functions, $\phi_k$, and the interaction
            functions $u_{km}$. For the single particle functions we use

@@

                \nabla_k\phi_k

@@

                \phi_k.

@@

            Note that the subscripts on the unit vectors $\vf{e}_i$ are

@@

                \nabla_k^2\phi_k

@@

@@

                \phi_k.
